---
title: "What makes a winning YC startup?"
subtitle: "And has the formula changed?"
author: "Alibek Mamyrbay &middot; Lawrence University Datathon"
output:
  revealjs::revealjs_presentation:
    theme: white
    highlight: pygments
    transition: slide
    self_contained: true
    css: theme.css
    reveal_options:
      slideNumber: true
      progress: true
      controls: true
      hash: true
      width: 1280
      height: 740
      margin: 0.06
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", out.width = "92%")
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(stringr)
  library(plotly); library(ggplot2); library(scales); library(forcats)
  library(reactable); library(htmltools); library(broom); library(ggrepel)
})
proj_root <- rprojroot::find_root_file(".", criterion = rprojroot::has_dir("data"))
knitr::opts_knit$set(root.dir = proj_root)
```

```{r load-data}
NOT_STARTED <- c("Summer 2026", "Fall 2026")
clean            <- read_csv(file.path(proj_root, "data/processed/yc_clean.csv"), show_col_types = FALSE) |>
  filter(!batch %in% NOT_STARTED)
q1_industry      <- read_csv(file.path(proj_root, "data/processed/q1_industry.csv"), show_col_types = FALSE)
q2_industry_year <- read_csv(file.path(proj_root, "data/processed/q2_industry_year.csv"), show_col_types = FALSE)
q3_era           <- read_csv(file.path(proj_root, "data/processed/q3_era_long.csv"), show_col_types = FALSE)
q1_logit         <- read_csv(file.path(proj_root, "data/processed/q1_logit.csv"), show_col_types = FALSE)
scrape_meta      <- jsonlite::fromJSON(file.path(proj_root, "data/raw/scrape_meta.json"))

YC_ORANGE <- "#F26625"
YC_DARK   <- "#1F2937"
GREY_MED  <- "#9CA3AF"
industry_palette <- c(
  "B2B"="#3B82F6","Consumer"="#F26625","Healthcare"="#10B981","Fintech"="#8B5CF6",
  "Industrials"="#F59E0B","Real Estate and Construction"="#EF4444",
  "Education"="#06B6D4","Government"="#6B7280"
)
```

## The dataset

<div class="two-col">
<div>
<h3>What we built</h3>
<ul>
```{r dataset-stats, include=FALSE}
analysis_batches <- setdiff(unique(clean$batch), "Unspecified")
analysis_years   <- clean |> filter(batch != "Unspecified") |> pull(batch_year) |> na.omit() |> unique()
```

<li>Scraped <b>`r format(nrow(clean), big.mark=",")`</b> companies across <b>`r length(analysis_batches)`</b> batches that have started (through Spring 2026)</li>
<li>Source: <a href="https://www.ycombinator.com/companies">ycombinator.com/companies</a> · <b>`r min(analysis_years)`–`r max(analysis_years)`</b>, `r length(analysis_years)` years</li>
<li>Hit Algolia search index directly (`YCCompany_production`)</li>
<li>Faceted by batch to bypass Algolia's 1,000-result pagination cap</li>
<li>Cleaned, joined, factored in R (`tidyverse` + `httr2`)</li>
</ul>
</div>
<div>
<h3>Status mix</h3>
```{r data-status, fig.width=5.2, fig.height=3.2, out.width="100%"}
clean |>
  count(status) |>
  filter(!is.na(status)) |>
  mutate(status = fct_reorder(status, n)) |>
  ggplot(aes(n, status, fill = status)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = scales::comma(n)), hjust = -0.15, color = YC_DARK, fontface = "bold") +
  scale_fill_manual(values = c("Active" = YC_ORANGE, "Acquired" = "#10B981",
                               "Public" = "#3B82F6", "Inactive" = "#9CA3AF")) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        axis.text.x = element_blank())
```
</div>
</div>

<p class="caption">Source: ycombinator.com/companies, scraped `r format(as.Date(scrape_meta$scrape_date), "%b %d, %Y")`. Data is YC's self-reported labels.</p>

## Three questions

<div class="big-questions">
<div class="q"><span class="qnum">1</span>What characteristics distinguish a <b>winning</b> YC startup?</div>
<div class="q"><span class="qnum">2</span>How has YC's <b>portfolio evolved</b> over 20 years?</div>
<div class="q"><span class="qnum">3</span>Has the <b>recipe for success</b> changed?</div>
</div>

<p class="caption">"Winner" = reached <b>Acquired</b> or <b>Public</b> status. "Mature" = batch ≥5 years old (companies have had time to exit).</p>

## Q1 · Fintech and B2B lead the exit hit rate {.section-q1}

```{r q1-industry, fig.width=11, fig.height=5.6, out.width="100%"}
q1_industry |>
  filter(group != "Unspecified") |>
  mutate(group = fct_reorder(group, p)) |>
  ggplot(aes(x = group, y = p)) +
  geom_segment(aes(xend = group, y = 0, yend = p), color = GREY_MED, linewidth = 0.5) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.25, color = YC_DARK, linewidth = 0.5) +
  geom_point(aes(size = n_total), color = YC_ORANGE) +
  geom_text(aes(y = hi, label = sprintf("%.0f%%  (n=%d)", p * 100, n_total)),
            hjust = -0.15, vjust = 0.5, size = 4.2, color = YC_DARK, fontface = "bold") +
  scale_y_continuous(labels = label_percent(), limits = c(0, 0.85), expand = expansion(mult = c(0, 0.05))) +
  scale_size_continuous(range = c(3, 10), guide = "none") +
  coord_flip() +
  labs(x = NULL, y = "Exit hit rate (Acquired ∪ Public among mature, non-Active companies)") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        axis.title.x = element_text(face = "bold"))
```

<p class="takeaway">Among mature batches: Fintech and B2B both ~52%. Consumer drags the average — startups serving everyday users are fundamentally harder.</p>

## Q1 · Survival is a function of size — but size is also an outcome {.section-q1}

<div class="two-col">
<div>
```{r q1-team, fig.width=6, fig.height=4.5, out.width="100%"}
read_csv(file.path(proj_root, "data/processed/q1_team.csv"), show_col_types = FALSE) |>
  mutate(group = factor(group, levels = c("1-5","6-10","11-25","26-100","101-500","500+"))) |>
  ggplot(aes(x = group, y = p)) +
  geom_col(fill = YC_ORANGE, alpha = 0.85, width = 0.6) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.2, color = YC_DARK) +
  geom_text(aes(label = sprintf("%.0f%%", p * 100)),
            vjust = -0.5, color = YC_DARK, fontface = "bold", size = 4,
            position = position_nudge(y = 0.05)) +
  scale_y_continuous(labels = label_percent(), limits = c(0, 1.1), expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Team size (current)", y = "Exit hit rate") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor = element_blank())
```
</div>
<div class="logit-side">
<h3>Controlling for everything: <b>team size still wins</b></h3>
<p style="margin-top:6px">Logit of <b>Successful vs Failed</b> on industry, region, team-size bucket, and era:</p>
<ul>
<li>26–100 employees vs 1–5: <b>7.1×</b> higher odds (p&lt;0.001)</li>
<li>101–500 vs 1–5: <b>14.7×</b> (p&lt;0.001)</li>
<li>USA vs Non-US: <b>1.3×</b> (n.s.) — geography barely matters</li>
<li>2015–2021 vs 2005–2014: <b>0.71×</b> (p=0.013)</li>
</ul>
<p class="caption-sm">n=`r format(nobs(glm(I(outcome=="Successful") ~ 1, data=clean, family=binomial)), big.mark=",")`+. Reference: B2B / 1–5 / Non-US / 2005–2014.</p>
</div>
</div>

## Q2 · B2B has eaten YC. Consumer has lost the throne. {.section-q2}

```{r q2-evo, fig.width=11.5, fig.height=5.4, out.width="100%"}
industry_levels <- q2_industry_year |>
  group_by(industry) |> summarise(total = sum(n)) |> arrange(desc(total)) |> pull(industry)
industry_levels <- setdiff(industry_levels, "Unspecified")
stack_order <- rev(industry_levels)

evo <- q2_industry_year |>
  filter(industry != "Unspecified", batch_year >= 2007, batch_year <= 2024) |>
  mutate(tip = sprintf("Batch %d — %s\n%d companies (%.1f%% of batch)",
                       batch_year, industry, n, share * 100))

p <- plot_ly()
for (ind in stack_order) {
  d <- evo |> filter(industry == ind)
  p <- p |> add_trace(
    data = d, x = ~batch_year, y = ~share,
    name = ind, legendgroup = ind,
    type = "scatter", mode = "none",
    stackgroup = "one", groupnorm = "percent",
    fillcolor = unname(industry_palette[[ind]]),
    text = ~tip, hoverinfo = "text"
  )
}
p |> layout(
    xaxis = list(title = "<b>Batch year</b>", dtick = 2, gridcolor = "white"),
    yaxis = list(title = "<b>Share of batch</b>", ticksuffix = "%", gridcolor = "white"),
    legend = list(orientation = "v", x = 1.02, traceorder = "reversed"),
    paper_bgcolor = "white", plot_bgcolor = "#F9FAFB",
    margin = list(t = 20, l = 70, r = 30, b = 50),
    hovermode = "closest"
  ) |>
  config(displaylogo = FALSE,
         modeBarButtonsToRemove = c("autoScale2d","lasso2d","select2d"))
```

<p class="takeaway">2010: Consumer ~35% of batch. 2024: barely 10%. The post-mobile gold rush is over; B2B SaaS now defines YC.</p>

## Q2 · YC has gone global {.section-q2}

```{r q2-map, fig.width=11, fig.height=5.0, out.width="100%"}
iso_map <- c(
  "USA"="USA","United Kingdom"="GBR","Canada"="CAN","India"="IND","Mexico"="MEX",
  "France"="FRA","Germany"="DEU","Brazil"="BRA","Spain"="ESP","Argentina"="ARG",
  "Singapore"="SGP","Australia"="AUS","Netherlands"="NLD","Israel"="ISR","Japan"="JPN",
  "Sweden"="SWE","Switzerland"="CHE","Nigeria"="NGA","Kenya"="KEN","South Africa"="ZAF",
  "Egypt"="EGY","Indonesia"="IDN","Vietnam"="VNM","Pakistan"="PAK","Philippines"="PHL",
  "Colombia"="COL","Chile"="CHL","Peru"="PER","Estonia"="EST","Ireland"="IRL",
  "Italy"="ITA","Portugal"="PRT","Poland"="POL","Denmark"="DNK","Norway"="NOR",
  "Finland"="FIN","Belgium"="BEL","Turkey"="TUR","South Korea"="KOR","China"="CHN",
  "Hong Kong"="HKG","Taiwan"="TWN","UAE"="ARE","Saudi Arabia"="SAU","Bangladesh"="BGD",
  "New Zealand"="NZL","Czech Republic"="CZE","Romania"="ROU","Hungary"="HUN","Greece"="GRC",
  "Lithuania"="LTU","Latvia"="LVA","Bulgaria"="BGR","Croatia"="HRV","Slovenia"="SVN",
  "Ghana"="GHA","Uganda"="UGA","Tanzania"="TZA","Ethiopia"="ETH","Morocco"="MAR",
  "Tunisia"="TUN","Senegal"="SEN","Rwanda"="RWA","Ecuador"="ECU","Uruguay"="URY",
  "Bolivia"="BOL","Paraguay"="PRY","Costa Rica"="CRI","Panama"="PAN","Guatemala"="GTM",
  "Dominican Republic"="DOM","Iceland"="ISL","Luxembourg"="LUX","Sri Lanka"="LKA",
  "Slovakia"="SVK","Serbia"="SRB","Belarus"="BLR","Kazakhstan"="KAZ","Uzbekistan"="UZB",
  "Georgia"="GEO","Armenia"="ARM","Lebanon"="LBN","Mongolia"="MNG","Nepal"="NPL",
  "Malaysia"="MYS","Thailand"="THA","Austria"="AUT","Algeria"="DZA","Bahrain"="BHR",
  "Namibia"="NAM","Venezuela"="VEN","Zambia"="ZMB","Kyrgyzstan"="KGZ","Puerto Rico"="PRI",
  "London"="GBR","Bengaluru"="IND"
)
country_counts <- clean |>
  filter(!is.na(country), country != "Remote") |>
  count(country, name = "n_companies") |>
  mutate(iso3 = unname(iso_map[country])) |>
  filter(!is.na(iso3))

plot_geo(country_counts) |>
  add_trace(z = ~n_companies, locations = ~iso3, text = ~country,
            color = ~n_companies, colors = c("#FFE4D6", "#F26625", "#7C2D12"),
            hovertemplate = "<b>%{text}</b><br>%{z} YC companies<extra></extra>",
            marker = list(line = list(color = "white", width = 0.5))) |>
  colorbar(title = "Companies", thickness = 12, len = 0.55) |>
  layout(geo = list(showframe = FALSE, projection = list(type = "natural earth"),
                    showcountries = TRUE, countrycolor = "#E5E7EB", bgcolor = "white"),
         margin = list(t = 0, l = 0, r = 0, b = 0)) |>
  config(displaylogo = FALSE)
```

<p class="takeaway">Top non-US: India (190), UK (161), Canada (133), Mexico (79). Once a Bay Area club, YC is now a global accelerator.</p>

## Q3 · The 2010s playbook stopped paying off {.section-q3}

```{r q3-slope, fig.width=10, fig.height=5.4, out.width="92%"}
slope <- q3_era |>
  mutate(industry = factor(industry, levels = sort(unique(industry)))) |>
  select(industry, era_q3, p, n_total, lo, hi)
slope_colors <- unname(industry_palette[levels(slope$industry)])

plot_ly() |>
  add_trace(
    data = slope,
    x = ~era_q3, y = ~p, color = ~industry, colors = slope_colors,
    type = "scatter", mode = "lines+markers",
    line = list(width = 4),
    marker = list(size = 13, line = list(color = "white", width = 2)),
    text = ~sprintf("<b>%s</b><br>%s: %.0f%% (n=%d)<br>95%% CI: %.0f–%.0f%%",
                    industry, era_q3, p*100, n_total, lo*100, hi*100),
    hoverinfo = "text"
  ) |>
  layout(
    xaxis = list(title = "<b>Batch era</b>", showgrid = FALSE),
    yaxis = list(title = "<b>Exit hit rate</b>", tickformat = ".0%", range = c(0.2, 0.65)),
    paper_bgcolor = "white", plot_bgcolor = "#F9FAFB",
    margin = list(t = 20, l = 70, r = 30, b = 50),
    hovermode = "closest", showlegend = TRUE
  ) |>
  config(displaylogo = FALSE)
```

<p class="takeaway">Every industry's hit rate dropped from 2005–2014 to 2015–2021. Consumer fell hardest (38% → 29%) — even controlling for batch maturity. The 2010s really were special.</p>

## Findings, in one slide {.dark-slide}

<div class="findings">
<div class="finding">
<h3 class="orange">1\. Industry matters less than you think</h3>
<p>Fintech (52%) ≈ B2B (52%). Consumer (34%) is the loser. After controlling for size & era, raw industry effects shrink.</p>
</div>
<div class="finding">
<h3 class="orange">2\. Geography barely matters</h3>
<p>Non-US success rate is <b>45%</b> vs USA <b>46%</b>. The "you must be in the Bay Area" myth is dead — at least for clearing the YC bar.</p>
</div>
<div class="finding">
<h3 class="orange">3\. The era effect is real</h3>
<p>Mature 2015–2021 batches exit at <b>0.71×</b> the rate of 2005–2014, controlling for everything else. The 2010s were a uniquely fertile decade.</p>
</div>
</div>

## Limitations & ethics

<div class="two-col">
<div>
<h3>Methodological caveats</h3>
<ul>
<li><b>Survivorship bias:</b> YC may de-emphasize Inactive companies — true failure rate could be higher</li>
<li><b>Censoring:</b> "exit hit rates" exclude still-active companies; recent batches still have time to surprise</li>
<li><b>Self-reported labels:</b> Status, industry, team size are YC's classification, not audited</li>
<li><b>500+ team bucket:</b> showed perfect separation in the regression (100% success) — excluded from the model</li>
</ul>
</div>
<div>
<h3>Sustainability & ethics</h3>
<ul>
<li>Public Algolia API only — no scraping behind auth walls</li>
<li>1 request/second rate limit, identifying User-Agent</li>
<li>No PII beyond what YC already publishes</li>
<li>Dataset published alongside this deck for verification & reuse</li>
<li>Source: <a href="https://www.ycombinator.com/companies">ycombinator.com/companies</a> (scraped `r format(as.Date(scrape_meta$scrape_date), "%b %d, %Y")`)</li>
</ul>
</div>
</div>

## Thank you {.dark-slide .closing}

<div class="closing-content">
<h1 class="orange">Questions?</h1>
<p class="byline">Code + dataset + this deck shipped together.</p>
<p class="caption-sm" style="color:#9CA3AF;margin-top:32px">
Built in R with <code>httr2</code>, <code>tidyverse</code>, <code>ggplot2</code>, <code>plotly</code>, <code>reactable</code>, <code>revealjs</code>.
</p>
</div>
